Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-13 main.Rmd
minor modifications on spelling and grammar

The remainder of the paper is organized as follows: Section \ref{sec:linear-time} provide some background material on linear granularities and introduces calendar algebra for computing different linear granularities. Section \ref{sec:cyclic-gran} formally characterizes different cyclic time granularities by extending the framework of linear time granularities. Section \ref{sec:cyclic-calendar} introduces cyclic calendar algebra for computing cyclic time granularities. Section \ref{sec:data-structure} discusses the data structure for exploring the conditional distributions of the associated time series across pairs of cyclic time granularities. Section \ref{sec:visualization} discusses the role of different factors in constructing an informative and trustworthy visualization. Section \ref{sec:application} examines how systematic exploration can be carried out for a temporal and non-temporal application. Section \ref{sec:discussion} summarizes this paper and discusses possible future direction.

@@

For example, $day \trianglelefteq week$ and $day \preceq week$ since every granule of "week" is the union of some set of granules of "day" and each "day" is a subset of a "week". Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions.

@@

\noindent For example, in a Gregorian calendar without leap years we could say days group periodically into months with the period $P = 365$ and the number of granules of "months" in each period given by $R=12$. In a Gregorian calendar with leap years, days group quasi-periodically into months with the <!--The value of $P$ and $R$ would still be 365 and 12 respectively, while the $E_i$s--> exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.

@@

**Example**: Example showing quasi-circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:quasi-circular-example}. Each granule is represented by a box. Two linear granularities $G' = Alter^{2}_{(1,-1)}(BG, Group_3(B))$ and $H' = Alter^{2}_{(1,-2)}(BG, Group_7(B))$ are considered. <!--This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different.--> $Q_{B, G'}$ and $Q_{B, H'}$ are repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$ or $H'$. For $G'$, $T_0 = \{0,1\}$ and $T_1 = \{2, 3, 4\}$. For $H'$, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$, then we will have \autoref{eq:eq4} and  \autoref{eq:circular-gran}.

@@

The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple:** and (2) **multiple-to-single:** which entails the representation of multiple-order-up cyclic granularities from single order-up cyclic granularities and vice versa. The hierarchical structure of time creates a natural nested ordering which can produce _single-order-up_ or _multiple-order-up_ granularities. We shall use the notion of a hierarchy table and order to define them.

@@

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ such that $C_i$ maps index set to $n_i$ levels or categories and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities each relating one finer linear granularity with a coarser one could be constructed. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, this big problem is broken down by focusing on two cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularity does.

@@

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity correspond to multiple values of the measured variable. This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. But that approach  hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of these multiple observations across cyclic granularities using some summary statistics, each of which can highlight different features of the data or through probability densities could be a potential way to compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

@@

Even after screening harmonies from the list of all possible combinations of cyclic time granularities that could be plotted together, the options could be too many and it might be useful to rank these harmonies. Ranking could be based on how well a harmony pair displays the maximum variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most important display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. And the harmony pair which has more variation between x-axis levels for all facet levels on an average is likely to be a good measure of how important a harmony pair is displaying the insights of the data. Distances between probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leiber divergence measure. Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth.
<!-- Suppose we have ${x_1, x_2, \dots, x_n} which are random sample from some underlying population. --> <!-- Let $p(x)$ and $q(x)$ be two probability distributions of a random variable x. -->
<!-- We compute Kullback-Liebler distance based on quantiles $q_1$ and $q_2$ of $v$ at probabilities $p$ corresponding to $C_1$ and $C_2$ respectively. -->

@@

3. Pairwise distances are computed between different quantiles of the time series variables using Jensen-Shanon distance. suppose $q$ is the sample quantile vector computed for probability vector $p$. Then Jensen-Shanon distance between two sample quantiles $q_j$ and $q_k$ will be given by:

$$d_{jk} = [D(q_j, r) + D(q_k, r)]/2 \quad where \quad r = (q_j + q_k)/2$$
where,
 $$D(q_j,q_k) = \int^{\infty}_{-\infty}q_j(x)logq_j(x)/q_k(x)\,dx$$
is the Kullback-Leibler divergence of $q_j$ with respect to $q_k$.

4. For each $\forall i = {1, 2, \dots, l}$, the maximum pairwise distance could be chosen from  ${m \choose 2}$ distances using $D_{j} = max(d_{jk})/(max(d_{jk}) - min(d_{jk}))$ $\forall j, k = {1, 2, \dots, m}$. Here, maximum is being adjusted against the range of the data to accommodate for comparison of maximum distances for different ranges.

5. The mean maximum pairwise distance (MMPD) over l = {1, 2, \dots, l}$ is computed by 
$MMPD = (1/l)\sum_{1}^{l}{D_{j}}$.

@@

## Ranking harmonies based on total variability

Even after screening harmonies from the list of all possible combinations of cyclic time granularities that could be plotted together, the options could be too many and it might be useful to rank these harmonies. Ranking could be based on how well a harmony pair displays the maximum variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most important display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. And the harmony pair which has more variation between x-axis levels for all facet levels on an average is likely to be a good measure of how important a harmony pair is displaying the insights of the data. Distances between probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leiber divergence measure. Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth.

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Mon ash University for sharing their wisdom and experience of developing R packages, Dr. Peter Tosca [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). We would also like to thank Nicholas Prison for many useful discussions and feedback on the manuscript. More details about the package 
can be found on the package website https://sayani07.github.io/gravitas/. This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article and the code is also available online
in the supplemental materials.

>>>>

The remainder of the paper is organized as follows: Section \ref{sec:linear-time} provides some background material on linear granularities and introduces calendar algebra for computing different linear granularities. Section \ref{sec:cyclic-gran} formally characterizes different cyclic time granularities by extending the framework of linear time granularities. Section \ref{sec:cyclic-calendar} introduces cyclic calendar algebra for computing cyclic time granularities. Section \ref{sec:data-structure} discusses the data structure for exploring the conditional distributions of the associated time series across pairs of cyclic time granularities. Section \ref{sec:visualization} discusses the role of different factors in constructing an informative and trustworthy visualization. Section \ref{sec:application} examines how systematic exploration can be carried out for a temporal and non-temporal application. Section \ref{sec:discussion} summarizes this paper and discusses possible future direction.

@@

For example, $day \trianglelefteq week$ and $day \preceq week$ since every granule of $week$ is the union of some set of granules of $day$ and each $day$ is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions.

@@

\noindent For example, in a Gregorian calendar without leap years we could say $day$ groups periodically into $month$ with the period $P = 365$ and the number of granules of $month$ in each period given by $R=12$. In a Gregorian calendar with leap years, $day$ groups quasi-periodically into $month$ with the <!--The value of $P$ and $R$ would still be 365 and 12 respectively, while the $E_i$s--> exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.

@@

**Example**: 
<!-- Example showing quasi-circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:quasi-circular-example}. Each granule is represented by a box.  -->

Consider two linear granularities $G' = Alter^{2}_{(1,-1)}(BG, Group_3(B))$ and $H' = Alter^{2}_{(1,-2)}(BG, Group_7(B))$.<!--This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different.--> Then $Q_{B, G'}$ and $Q_{B, H'}$ are repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$ or $H'$. For $G'$, $T_0 = \{0,1\}$ and $T_1 = \{2, 3, 4\}$. For $H'$, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$, then we will have \autoref{eq:eq4} and  \autoref{eq:eq:eq5}.

@@

The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple** and (2) **multiple-to-single** which entails the representation of multiple-order-up cyclic granularities from single order-up cyclic granularities and vice versa. The hierarchical structure of time creates a natural nested ordering which can produce _single-order-up_ or _multiple-order-up_ granularities. We shall use the notion of a hierarchy table and order to define them.

@@

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ such that $C_i$ maps index set to $n_i$ levels or categories and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, this big problem is broken down by focusing on two cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the choice of the granularities does.

@@

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. But that approach  hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of these multiple observations across cyclic granularities using some summary statistics, each of which can highlight different features of the data or through probability densities could be a potential way to compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

@@

<!-- ## Choosing harmonies to show -->

<!-- Even after screening harmonies from the list of all possible pairs of cyclic time granularities, the options could be too many and it might be useful to rank these harmonies. Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most important display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. And the harmony pair which has more variation between x-axis levels for all facet levels on an average is likely to be a good measure of how important a harmony pair is displaying the insights of the data. Distances between probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leibler divergence measure. Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth. -->
<!-- Suppose we have ${x_1, x_2, \dots, x_n} which are random sample from some underlying population. --> <!-- Let $p(x)$ and $q(x)$ be two probability distributions of a random variable x. -->
<!-- We compute Kullback-Liebler distance based on quantiles $q_1$ and $q_2$ of $v$ at probabilities $p$ corresponding to $C_1$ and $C_2$ respectively. -->

@@

@@

Even after screening harmonies from the list of all possible combinations of cyclic time granularities that could be plotted together, the options could be too many and it might be useful to rank these harmonies. Ranking could be based on how well a harmony pair displays the maximum variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most important display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. And the harmony pair which has more variation between x-axis levels for all facet levels on an average is likely to be a good measure of how important a harmony pair is displaying the insights of the data. Distances between probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leibler divergence measure. Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth.

3. Pairwise distances are computed between different quantiles of the time series variables using Jensen-Shanon distance. Suppose $q$ is the sample quantile vector computed for probability vector $p$. Then Jensen-Shanon distance between two sample quantiles $q_j$ and $q_k$ will be given by:

$$d_{jk} = [D(q_j, r) + D(q_k, r)]/2 \quad where \quad r = (q_j + q_k)/2$$
where,
 $$D(q_j,q_k) = \int^{\infty}_{-\infty}q_j(x)log\frac{q_j(x)}{q_k(x)}\,dx$$
is the Kullback-Leibler divergence of $q_j$ with respect to $q_k$.

4. For each $\forall i = {1, 2, \dots, l}$, the maximum pairwise distance could be chosen from  ${m \choose 2}$ distances using $D_{i} = max(d_{jk})/(max(d_{jk}) - min(d_{jk}))$ $\forall j, k = {1, 2, \dots, m}$. Here, maximum is being adjusted against the range of the data to accommodate for comparison of maximum distances for different ranges.

5. The mean maximum pairwise distance (MMPD) over $l = {1, 2, \dots, l}$ is computed by 
$MMPD = (1/l)\sum_{i=1}^{l}{D_{j}}$.

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Mon ash University for sharing their wisdom and experience of developing R packages, Dr. Peter Tosca [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). We would also like to thank Nicholas Spyrison for many useful discussions and feedback on the manuscript. More details about the package 
can be found on the package website https://sayani07.github.io/gravitas/. This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article and the code is also available online
in the supplemental materials.
